# Exploring "sharkresignts.csv"

```{r}
bite_data <- read.csv("data/sharkresights.csv")

# View(bite_data)
summary(bite_data)

unique(bite_data$area)
length(unique(bite_data$area))

unique(bite_data$obssex)
unique(bite_data$obsage)
length(unique(bite_data$resightID)) # 225 resight ids
length(unique(bite_data$animalID))  #106 animals

unique(bite_data$markobs)
unique(bite_data$markobspos)

unique(bite_data$pctmolt)
unique(bite_data$area)

dim(bite_data)
sort(unique(bite_data$season))
```

## Columns Explained

All these rows are elephant seals that hang out at Ano Nuevo State Park.

-   resightID:

    -   All of the rows have unique resightIDs

-   animalID:

    -   ID of the animal observed

    -   numerical and repeats

-   season:

    -   numerical; the year the data was recorded (XXXX)

-   date:

    -   char; MM/DD/YY

-   area:

    -   char
    -   Abbreviations of the area the seal was found

-   regionID:

    -   numerical, discrete

    -   range from 1-17; the numerical "representation" of the areas

-   markobs:

    -   char

    -   A mark exists on the animal at the time of observation

    -   Lots of missing data, don't know what it means

-   markobspos:

    -   char

    -   Also lots of missing data,

    -   Location of the marks

    -   L, R Lsh, idk what it means

-   markadded:

    -   char

    -   When a mark was added to the seal

    -   Similar problem to markobs

-   markaddedpos

    -   char

    -   Where the mark was added

    -   similar problems and values to markobspos

-   obssex:

    -   char

    -   M/F

    -   Misssing values

-   obsage:

    -   char

    -   Values: Y, ad, juv, SA4, a-bY

    -   some missing/empty data

-   pctmolt:

    -   numerical

    -   Doesn't relate to us; seals literally molt their skin and hair; animal need to be marked once they finish molting (mark comes off with molting)

    -   Lots of missing values

-   comment:

    -   characters

    -   Literall comments of various sizes

-   markcomment:

    -   char

    -   comment on comment

-   dead:

    -   numerical

    -   0/1, technically binary (1 = observation was dead)

    -   No missing values

-   observer

    -   char

    -   names: X. XXX

-   entrydate:

    -   char

    -   MM/DD/YY

-   entry:

    -   char

    -   names: X. XXX

    -   Aren't always the same as the observer!

-   estimated:

    -   num, literally all 0's, no missing values

-   shark

    -   char

    -   "Comment includes shark" for all rows

-   unsure:

    -   num

    -   Missing lots of values (180)

    -   Binary data (0/1)

-   cookiecutter:

    -   numerical; technically binary

    -   1 means cookie_cutter shark bite

    -   Missing lots of values (180)

## Gender of Seals grouped by year

```{r}
# Number of males/females logged by decade
sum(bite_data$obssex %in% c("F", "f") & bite_data$season %in% c("1977", "1979"))
sum(bite_data$obssex %in% c("M") & bite_data$season %in% c("1977", "1979"))
sum(bite_data$obssex %in% c("", NA) & bite_data$season %in% c("1977", "1979"))

sum(bite_data$obssex %in% c("F", "f") & bite_data$season %in% c("1982", "1986", "1988", "1989"))
sum(bite_data$obssex %in% c("M") & bite_data$season %in% c("1982", "1986", "1988", "1989"))
sum(bite_data$obssex %in% c("", NA) & bite_data$season %in% c("1982", "1986", "1988", "1989"))

sum(bite_data$obssex %in% c("F", "f") & bite_data$season %in% c("1990", "1991", "1993", "1995"))
sum(bite_data$obssex %in% c("M") & bite_data$season %in% c("1990", "1991", "1993", "1995"))
sum(bite_data$obssex %in% c("", NA) & bite_data$season %in% c("1990", "1991", "1993", "1995"))

c("2000":"2009")
sum(bite_data$obssex %in% c("F", "f") & bite_data$season %in% c("2002":"2009"))
sum(bite_data$obssex %in% c("M") & bite_data$season %in% c("2002":"2009"))
sum(bite_data$obssex %in% c("", NA) & bite_data$season %in% c("2002":"2009"))

# Attempt 2
library(dplyr)

bite_summary <- bite_data %>% count(season, obssex, name="total") %>% arrange(season, desc(total))
print(bite_summary)

bites_by_decade <- bite_summary %>% mutate(year = as.numeric(season)) %>% mutate(decade = floor(year / 10) * 10) %>% count(decade, obssex, name = "total") %>% arrange(decade, desc(total))
print(bites_by_decade)
```