# Shark Resight Data EDA

**Import Files and Make Dataframe**

```{r}
# File Set Up
## Import packages
install.packages("dplyr")
library(tidyverse)
library(stringr)
library(ggplot2)

## Import file (Sharry) (I have mine in "data")
sharkbite <- read.csv("data/sharkresights.csv") #access limited to research collaborators 

## Import file (Nat)
#setwd("/Users/nataliebilal/Library/CloudStorage/OneDrive-Personal/R practice") 

#sharkbite <- read.csv("sharkresights.csv") #access limited to research collaborators 

#selecting relevant variables
sharkbite_variables <- sharkbite %>% 
  select(animalID, season, date, area, regionID, markobs, obssex, obsage, comment, dead) 
View(sharkbite_variables) 
```

**Data Cleaning**

```{r}
#remove cookie cutter false positives
sharkbite_nocookie <- sharkbite_variables %>% 
  filter(!str_detect(comment, 'cutter')) %>%
  filter(!str_detect(comment, 'CC')) %>%
  filter(!str_detect(comment, 'not shark'))

#cleaning sex variable 
sharkbite_nocookie <- sharkbite_nocookie %>%
  mutate(obssex = case_when(
    obssex == "" ~ NA_character_,
    str_detect(obssex, 'F|f') ~ "F",
    str_detect(obssex, 'M') ~ "M",
    TRUE ~ obssex))

#categorical bite location variable
sharkbite_nocookie <- sharkbite_nocookie %>% 
  mutate(bite_location = case_when(
      str_detect(comment, 'flip|RF|FF') ~ "flipper",
      str_detect(comment, 'low|end|butt|rump|chin|eye|neck|face|mouth') ~ "end",
      TRUE ~ "body"))
unique(sharkbite_nocookie$bite_location)

#categorical ageclass variable
sharkbite_nocookie <- sharkbite_nocookie %>% 
  mutate(age_binary = case_when(
      str_detect(obsage, 'ad|Ad') ~ "adult",
      obsage == "" ~ NA_character_,
      TRUE ~ "juvenile"))
 
#recode non-Año locations
sharkbite_nocookie <- sharkbite_nocookie %>%
  mutate(regionID = case_when(
    regionID >= 12 & regionID <= 18 ~ NA,
    TRUE ~ regionID))

#recode locations to be more specific 
sharkbite_nocookie <- sharkbite_nocookie %>%
  mutate(regionID = case_when(
    area %in% c("1", "17", "8B") ~ 1,
    area %in% c("TS", "TSB", "SBE", "SBEL") ~ 2,
    area %in% c("TSW", "TSC", "TSD") ~ 3,
    area %in% c("BBSL", "AP", "APB", "SB", "SBW") ~ 4,
    area == "APG" ~ 5,
    area %in% c("BMB", "BMb") ~ 6,
    area %in% c("BMN", "BMNN", "BMS", "BMC", "BBSU", "BM") ~ 7,
    area %in% c("MBB", "MBB/BBN", "MBBL") ~ 8,
    area == "MBBU" ~ 9,
    area %in% c("BBNS", "BBS") ~ 10,
    area == "BBN" ~ 11,
    area == "NPB1" ~ 12,
    area == "NP" ~ 13,
    area %in% c("NP0", "NP1") ~ 14,
    area == "NPG1" ~ 15,
    area %in% c("NP2", "npg2b") ~ 16,
    area %in% c("NPG3", "NPG4") ~ 17,
    area %in% c("NP3", "NP4", "NP5") ~ 18,
    TRUE ~ regionID))

#categorical North vs South vs Island areas of the colony 
sharkbite_nocookie <- sharkbite_nocookie %>%
  mutate(area_NSI = case_when(
    regionID == 1 ~ "island",
    regionID %in% c(2, 3, 4) ~ "south",
    regionID %in% c(13, 14, 15, 16, 17, 18, 19) ~ "north",
    regionID %in% c(5, 6, 7, 8, 9, 10, 11, 12) ~ NA_character_,
    TRUE ~ NA_character_))
unique(sharkbite_nocookie$area_NSI)

#categorical beach vs inland variable
sharkbite_nocookie <- sharkbite_nocookie %>% 
  mutate(area_beach = case_when(
       regionID %in% c(1, 2, 4, 6, 8, 10, 11, 12, 13, 15, 16, 17) ~ "beach",
       regionID %in% c(3, 5, 7, 9, 14, 18) ~ "inland",
    TRUE ~ NA_character_))
unique(sharkbite_nocookie$area_beach)

# "Final look at data"
summary(sharkbite_nocookie)
```

**All Unique Values of Columns**

```{r}
dim(sharkbite)
dim(sharkbite_nocookie)

unique(sharkbite_nocookie$animalID)
unique(sharkbite_nocookie$season)
# unique(sharkbite_nocookie$date)
# unique(sharkbite_nocookie$area)
unique(sharkbite_nocookie$regionID)
# unique(sharkbite_nocookie$markobs)
unique(sharkbite_nocookie$obssex)
# unique(sharkbite_nocookie$obsage)
# unique(sharkbite_nocookie$comment)
# unique(sharkbite_nocookie$dead)
unique(sharkbite_nocookie$bite_location)
unique(sharkbite_nocookie$age_binary)
unique(sharkbite_nocookie$area_NSI)
unique(sharkbite_nocookie$area_beach)

# Row 16 has an invalid date! May need to delete
# sharkbite_nocookie[16,]
```

**Make Unique Dataframe** Don't run this; it will give an error!

```{r}
# Chi-Square Test for Independence: Age Class by Sex

##Research question: what is the more common sex of older seals who return with sharkbites?
##Hypothesis: older seals are more likely to be female because this sex forages in lower risk waters for shark encounters

##set up: finding unique seals 
duplicate_seal |> add_count(animalID) |> filter(n > 1) |> arrange(animalID)

duplicate_seal_sorted <- duplicate_seal |> add_count(animalID) |> filter(n > 1) |> arrange(animalID)
duplicate_seal_sorted

#seals with 2 ages - need to just keep adult observations 
#   9615 14243 17692 41117 47026 49594 

#seal with 2 sexes (as a juvenile, no other sexes identified) - set as NA
#   55642

#arrange in recency order and keep only unique animalIDs
unique_seals <- sharkbite_nocookie |>
  group_by(animalID) |>
  slice_max(order_by = season, n = 1, with_ties = FALSE) |>
  ungroup()
unique_seals

##set up: creating 2x2 table
chi_table <- table(unique_seals$obssex, unique_seals$age_binary)
chi_table

# Analysis: Fisher's exact test for small samples using table 
fish_test <- fisher.test(matrix(c(23, 3, 5, 19), nrow = 2))

##interpretation: survivor shark-bitten juveniles are more likely to be male, and adults are more likely to be female 

#Visualization: create dataframe from contingency table for plotting 
##create dataframe 
chi_df <- as.data.frame(chi_table)
colnames(chi_df) <- c("Sex", "Age Class", "Count")
chi_df

##rearrange for easy plotting 
chi_df_plot <- chi_df[rep(1:nrow(chi_df), chi_df$Count), 1:2]
rownames(chi_df_plot) <- NULL
chi_df_plot

##plot
library(ggstatsplot)
chi_plot <- ggbarstats(data = chi_df_plot,
  x = Sex,
  y = `Age Class`,         
  results.subtitle = FALSE,
  subtitle = paste0(
    "Fisher's exact test, p-value ",
    ifelse(fish_test$p.value < 0.001, "< 0.001", round(fish_test$p.value, 3))))
chi_plot

#I'll leave colors up to you if you'd like!
```

**Bar Graphs for Each Variable** *(age class, area, bite location, sex, year)*

```{r}
summary(sharkbite_nocookie)

# Make binary columns based on bite location
sharkbite_nocookie$flipper_bite <- ifelse(sharkbite_nocookie$bite_location == "flipper", 1, 0)
sharkbite_nocookie$headrear_bite <- ifelse(sharkbite_nocookie$bite_location == "end", 1, 0)
sharkbite_nocookie$body_bite <- ifelse(sharkbite_nocookie$bite_location == "body", 1, 0)
unique(sharkbite_nocookie$obssex)

# Make regionID a factor column not a numeric column:
sharkbite_nocookie$regionID <- as.factor(sharkbite_nocookie$regionID)
sharkbite_nocookie$flipper_bite <- as.factor(sharkbite_nocookie$flipper_bite)
sharkbite_nocookie$headrear_bite <- as.factor(sharkbite_nocookie$headrear_bite)
sharkbite_nocookie$body_bite <- as.factor(sharkbite_nocookie$body_bite)


# GRAPHS
# Age Class
ggplot(sharkbite_nocookie, aes(x = age_binary)) + geom_bar() + labs(x = "Age (adult or juvinille)", y = "Count", title = "Count by Age")

# Area
ggplot(sharkbite_nocookie, aes(x = factor(regionID), fill = area_beach)) + 
  geom_bar() + 
  geom_text(stat = "count", aes(label = after_stat(count)), hjust = -0.2) + 
  coord_flip() +
  labs(x = "Region", y = "Count", fill = "Location", title = "Sighting by Region and Inland vs Beach") +
  theme_minimal()

# Bite Location
ggplot(sharkbite_nocookie, aes(x = bite_location, fill = obssex)) +
  geom_bar(position = "stack") +
  labs(x = "Bite Location", y = "Count", fill = "Gender", title = "Total Bites by Location and Gender") +
  theme_minimal()

# Sex
ggplot(sharkbite_nocookie, aes(x = obssex)) + geom_bar() + labs(x = "Seal's gender", y = "Count", title = "Count by Gender")

# Year
ggplot(sharkbite_nocookie, aes(x = season)) + geom_bar() + labs(x = "Year of Obsevations", y = "Count", title = "Number of observations by Year")

```

**Logistic Regression attempt:**

-   flipper_log: Prediction of getting a Flipper bite by gender and age

-   headrear_log: Prediction of getting a Head/rear bite by gender and age

-   body_log: Prediction of getting a Body bite by gender and age

```{r}
# Clean data One last time
sharkbite_log <- sharkbite_nocookie[complete.cases(sharkbite_nocookie$obssex), ]
dim(sharkbite_log)
dim(sharkbite_nocookie)

# Fit the logistic model
flipper_log <- glm(flipper_bite ~ obssex * age_binary, data = sharkbite_log, family = binomial(link = 'logit'))
headrear_log <- glm(headrear_bite ~ obssex * age_binary, data = sharkbite_log, family = binomial(link = 'logit'))
body_log <- glm(body_bite ~ obssex * age_binary, data = sharkbite_log, family = binomial(link = 'logit'))

# View results
summary(flipper_log)
summary(headrear_log)
summary(body_log)
```

**Checking Assumptions**

```{r}
# Check graphs for log regression
par(mfrow = c(2, 2))
plot(flipper_log)
plot(headrear_log)
plot(body_log)

par(mfrow = c(1, 1))
```